---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2026 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2026`- `Fall 2021`)/`Fall 2021` * 100)
```



### Task 4: Describe the results

Write a few sentences about what you found in those percentages. What's the story here? What questions do you have?

The most significant percent change is under student sustainability with a 150% change. The second category with the biggest percent change was for the shuttle buses with a 58.30% change. I think that the most relevant topic to write about would be the shuttle buses because the school made a switch to electric buses, which could fall under either the student sustainability or shuttle bus category because it involves sustainability initiatives. I would question how the electric buses may have reshaped the fees and how much of the project was covered by grants and how much of the costs students absorbed. 



When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
