---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE
A part of the data that stood out to me was that there are 10 employees that have a salary listed as 0. On the other end, there are 9 detailees and part time detailees that are making 197200. With this information, I would be curious about the disparity between employee and detailee salaries. The ten 0 salaries would cause problems during analysis. 
### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries %>% 
  summarize(
    mean_salary=mean(SALARY,na.rm=TRUE),
    median_salary=median(SALARY,na.rm=TRUE)
  )
```


### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries %>% 
  summarize(
    mean_salary=mean(SALARY,na.rm =TRUE),
    median_salary=median(SALARY,na.rm =TRUE),
    min_salary=min(SALARY,na.rm=TRUE),
    max_salary=max(SALARY,na.rm=TRUE)
  )
  
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE
The mean(112,702.70) and median(109,243) are fairly close to one another, so White House salaries are likely to be evenly distributed. The Min and Max salaries show that there is a huge gap between the top earners and that the bottom earners. The outliers in the data are the employees that have their salaries listed as 0

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
Based on the data, I would investigate why there are ten people with an annual salary listed as $0. Listing their salary as $0 brings to question if there are employees truly doing unpaid work or if it hides something greater like them being payed through different routes for tax evasion or if they are getting paid significantly more than others and what the reason is for that. 
