---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
#installed.packages("janitor")
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities<-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(
    violent_rate_2019 = (violent2019 / pop2019) * 1000,
    property_rate_2019 = (property2019 / pop2019) * 1000,
    violent_rate_2020 = (violent2020 / pop2020) * 1000,
    property_rate_2020 = (property2020 / pop2020) * 1000
  )
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE
Baltimore has the highest rate of violent crime in 2019 and Ocean City has the highest rate of property crime in 2019. In 2020 Ocean City has the highest rate of both violent and property crimes. These cities are both very high traffic and tourist attractions, which is a predictable pattern since they are both two of the most popular major cities in Maryland. As someone who did not grow up in Maryland and does not know the state well, I would not have predicted that Ocean City had the highest rate of property and violent crime in 2020. A story could be that due to Covid-19, the rate of crime in Baltimore decreased so sharply that it was no longer the city with the highest rate of crime in the state. With more than just this dataset, I would examine how trends of crime in Baltimore changed after Covid-19 and if they got worse or better than before the pandemic happened. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change<-md_cities_rates|>mutate(violent_change=((violent_rate_2020-violent_rate_2019)/violent_rate_2019)*100,property_change= ((property_rate_2020-property_rate_2019)/property_rate_2019)*100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE
There are 6 cities that say there is infinite percent change for violent crimes and 4 that say infinite for percent change for property crimes. These outliers are interesting because three of the cities that say infinite percent change for property crimes also had no change in violent crimes. 


### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE
Ocean City had Maryland's highest violent and property crime rates in 2020, after Baltimore's crime rates dropped enough to lose its position as the state's highest-rate city. 

The change comes as Baltimore's crime rates fell sharply from 2019 to 2020, while Ocean City's rates increased. This shift raises questions about how the COVID-19 pandemic shaped crime patterns in Maryland's cities. 




### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: 
### 10a: Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile=ntile(violent_rate_2020,100))
```
### 10b: Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` 
```{r}
md_cities_percentiles |> summarise(cities_above_90th=sum(percentile>90))
```
### 10c:R's `IQR()` function. The syntax is similar to how `mean()` works.
```{r}
md_cities_rates |> summarise(IQR_violent_crime_2020= IQR(violent_rate_2020))
```
### 10d: Do the results change your best story idea from the previous assignment? Why or why not?
These results do not change my previous story idea. The 2020 violent crime rates show that there are 10 cities falling above the 90th percentile. The IQR of 4.198895 also shows that the middle 50% of cities had a relatively small range of violent crime rats, so I would still want to focus on cities with especially high crime rates compared to the rest of the state. 


