---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```
### Task 3: Sample 5 rows of data from the dataset
```{r}
sample5 <- mandms|> sample_n(5)
```

### Task 4: Calculate the mean age and standard deviation of the sample, using either the red or black number of M&Ms
```{r}
sample5 |> summarise(mean_red=mean(red), sd_red=sd(red))
```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
mandms|> summarise(mean_red=mean(red), sd_red=sd(red))
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The sample mean was 77.2 and the population mean was 72.98. The sample standard deviation was 22.28 and the population standard deviation was 16.46. The means are fairly similar, but the standard deviations have a larger difference. This shows that the sample was reasonably accurate and aligned with the population, but it did not match it exactly.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?
The results show that the average number of red M&Ms in the sample was close to the average for the full population, but the sample had more variation than the population. This shows that random sampling can produce results that are similar to the population, but small samples can still differ from the population.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
